Automatically Proving Equivalence by Type-Safe Reflection

### 2017-05-19 paper-CICM-17/content/1-introduction.tex
Final changes for me

Proofs assistants like Coq~\cite{BertotC04} 
and programming languages like 
Agda~\cite{DBLP:conf/tldi/Norell09} and
Idris~\cite{brady2013idris} are based on Intensional Type Theories that contain
two notions of equality: propositional equality, that can be
manipulated in the language, and judgemental (or definitional) equality.
Propositional equality corresponds to the mathematical notion: this is
a proposition that can be assumed, negated, proved or disproved. Since in type
theory, propositions are types~\cite{How80}, the proposition that two
elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id} (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

Proving that one term is equal to another is common in formal verification, and
proof obligations arise naturally in dependently typed programming when
indexing types over values in order to capture some logical properties.
To demonstrate this,
we revisit an example from previous work~\cite{DBLP:conf/plpv/Brady13} which
shows how proof obligations arise when a type is indexed by natural
numbers. Our goal is to implement a verified
library of binary numbers. To ensure functional correctness,
we define types \code{Bit} and \code{Binary},
indexed over the value they represent (expressed as a natural number):

@@

Our handwritten proof uses only the existence of a neutral
element, and the associativity and commutativity of $+$ on \code{Nat}. Thus,
we're rewriting a term by using the properties of a commutative monoid. We
would like a generic prover for commutative monoids to find a proof 
automatically.

>>>>

Proofs assistants like Coq~\cite{BertotC04} 
and programming languages like 
Agda~\cite{DBLP:conf/tldi/Norell09} and
Idris~\cite{brady2013idris} are based on Intensional Type Theories that contain
two notions of equality: propositional equality, that can be
manipulated in the language, and judgemental (or definitional) equality.
Propositional equality corresponds to the mathematical notion: this is
a proposition that can be assumed, negated, proved or disproved. Since in type
theory, propositions are types~\cite{How80}, the proposition that two
elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id}\ (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

Proving that one term is equal to another is common in formal verification, and
proof obligations arise naturally in dependently typed programming when
indexing types over values in order to capture some logical properties.
To demonstrate this,
we revisit an example from previous work~\cite{DBLP:conf/plpv/Brady13} which
shows how proof obligations arise when a type is indexed by natural
numbers. Our goal is to implement a verified
library of binary numbers. To ensure functional correctness,
we define the types \code{Bit} and \code{Binary} indexed over the value they represent (expressed as a natural number):

@@

Our handwritten proof \code{adc\_lemma\_2} uses only the existence of a neutral
element, and the associativity and commutativity of $+$ on \code{Nat}. Thus,
we're rewriting a term by using the properties of a commutative monoid. We
would like a generic prover for commutative monoids to find a proof 
automatically.

### 2017-05-19 paper-CICM-17/content/3-hierarchyOfTactics.tex
Final changes for me

We describe operations, constants and properties of each algebraic structure
in an interface. The first structure is
the magma. which is a structure built on top of \code{Set} that adds
\code{Plus} operation, and no specific properties:

@@

This code means that a type \code{c} (for $carrier$) is a \code{Magma} if it is
already a \code{Set} (ie, it is equipped with the equivalence relation $\simeq$
and the equivalence test \code{set\_eq}), and if it has a $+$ operation.  In
fact, there is an additional requirement that will apply to all operators (in
this case, the $+$ operation), which is that they need to be ``compatible" with
the equivalence relation, which is expressed by the following axiom for $+$:
\\ $\code{Plus\_preserves\_equiv} : \{c:\code{Type}\} \rightarrow
\{\code{Magma}\ c\} \rightarrow\{c1:c\} \rightarrow \{c2:c\} \rightarrow
\{c1':c\} \rightarrow \{c2':c\} \rightarrow (c1 \simeq c1') \rightarrow (c2
\simeq c2') \rightarrow ((c1+c2) \simeq (c1'+c2'))$

@@

When proving an equivalence $x \simeq y$, the universally-quantified
variables are abstracted and they become part of the context. Our
tatics normalise both sides of the ``potential equivalence'' $x
\simeq y$, and compare the results by syntactic equality. The
difficulty is that the normalisation function needs to consider
variables, constants and operators. For this
reason, we work by reflection, which allows us to work on syntax instead of values. We define one type of reflected
terms for each algebraic structure. The novelty is not the use of reflected
terms, but the use of a type-safe reflection
mechanism where we index the reflected terms by the concrete value that they
represent. Each of these datatype is parametrised over a type \code{c}, which
is the type on which we want to prove equalities (the \emph{carrier}
type). It is also indexed over an implementation of the corresponding interface
for \code{c} (we usually call it \code{p}, because it behaves as a \emph{proof}
telling that the structure \code{c} has the desired properties), indexed over a
context of abstracted variables (a vector $\Gamma$ of \code{n} elements of type
\code{c}). Most importantly, it is indexed over a value of type \code{c},
which is the concrete value being encoded.

@@

For an expression $e_x\ :\ \code{ExprMa}\ \Gamma\ x$, we say that $e_x$
denotes the value $x$ in the context $\Gamma$.  When an expression
is a variable \code{(VarMa\ \_\ $\Gamma$\ i)}, the denoted value is the
corresponding variable in the context, i.e., \code{(index\ i\ $\Gamma$)}.  The
expression \code{(ConstMa\ \_\ $\Gamma$\ k)} denotes the constant $k$ of type
$c$. Finally, if $e_x$ is an expression encoding the concrete value $x$, and
$e_y$ is an expression encoding the concrete value $y$, then the expression
\code{PlusMa\ $e_x$\ $e_y$} denotes the concrete value $(x + y)$. Because the
reflected terms embed their corresponding inputs, they are guaranteed to be
sound representations.

@@

We will explain how to implement the tactic for \code{Group} specifically,
and the other structures are implemented similarly.
The equivalence we are trying to prove is $x \simeq y$, where $x$ and $y$ are
elements of the type $c$, which implements a group. The reflected term for $x$
is denoted $e_x$, and has type $\code{ExprG}\ p\
\Gamma\ x$, which means that $e_x$ is guaranteed to be the encoding of $x$.  
Similarly,
$y$, is encoded by $e_y$, and its type is indexed by the
value $y$.  Evaluating \code{norm} on $e_x$ produces the
normal form $e_{x'}$ of type $\code{ExprG}\ p\ \Gamma\ x'$ and a proof $p_x$ of
$x \simeq x'$.
Similarly, for $e_y$, it produces
the normal form $e_{y'}$ of type $\code{ExprG}\ p\ \Gamma\ y'$ and a proof
$p_y$ of $y \simeq y'$.  It now suffices to compare the normal forms $e_{x'}$ and $e_{y'}$
using a syntactic equality test \code{exprG\_eq}. If the two normalised expressions $e_{x'}$ and $e_{y'}$ are equal, then
they necessarily have the same type\footnote{We work with the
heterogeneous equality by default in Idris, but as always, the only way to
have a proof of \code{a:A = b:B} is when \code{A = B}}, and therefore $x'=y'$, which implies
$x' \simeq y'$ because the propositional equality is more restrictive than an
equivalence relation.  By using the two equivalences $x \simeq x'$ and $y
\simeq y'$, and the new equivalence
$x' \simeq y'$, we can use \code{eq\_preserves\_eq}, to get the desired proof of
$x \simeq y$. We put all of this together in a function
\code{buildProofGroup}:

@@

Each algebraic structure is equipped with a function for reducing reflected terms to their normal form.
The algebraic theories which concern us admit a canonical representative\footnote{This only holds in the absence of additional axioms}  for
any element, a property which we use to decide equalities. Without this
property, it would be more complicated to decide equivalence without
brute-forcing a series of rewritings, that would have no termination guarantee.

>>>>

We describe operations, constants and properties of each algebraic structure 
in an interface. The first algebraic theory is \code{Magma}, which is a structure built on top of \code{Set} that adds
\code{Plus} operation, and no specific properties:

@@

This code means that a type \code{c} (for $carrier$) is a \code{Magma} if it is
already a \code{Set} (ie, it is equipped with the equivalence relation $\simeq$
and the equivalence test \code{set\_eq}), and if it has a $+$ operation.  In
fact, there is an additional requirement that will apply to all operations (in
this case, the $+$ operation), which is that they need to be ``compatible'' with
the equivalence relation, which is expressed by the following axiom for $+$:
\\ $\code{Plus\_preserves\_equiv} : \{c:\code{Type}\} \rightarrow
\{\code{Magma}\ c\} \rightarrow\{c1:c\} \rightarrow \{c2:c\} \rightarrow
\{c1':c\} \rightarrow \{c2':c\} \rightarrow (c1 \simeq c1') \rightarrow (c2
\simeq c2') \rightarrow ((c1+c2) \simeq (c1'+c2'))$

@@

When proving an equivalence $x \simeq y$, the universally-quantified
variables are abstracted and they become part of the context. Our
tatics normalise both sides of the ``potential equivalence'' $x
\simeq y$, and compare the results by syntactic equivalence. The
difficulty is that the normalisation function needs to consider
variables, constants and operators. For this
reason, we work by reflection, which allows us to work on syntax instead of values. We define one type of reflected
terms for each algebraic structure. The novelty is not the use of reflected
terms, but the use of a type-safe reflection
mechanism where we index the reflected terms by the concrete value that they
represent. Each of these datatype is parametrised over a type \code{c}, which
is the type on which we want to prove equalities (the \emph{carrier}
type). It is also indexed over an implementation of the corresponding interface
for \code{c} (we usually call it \code{p}, because it behaves as a \emph{proof}
telling that the structure \code{c} has the desired properties), indexed over a
context of abstracted variables (a vector $\Gamma$ of \code{n} elements of type
\code{c}). Most importantly, it is indexed over a value of type \code{c},
which is the concrete value being encoded.

@@

For an expression $e_x\ :\ \code{ExprMa}\ \Gamma\ x$, we say that $e_x$
denotes the value $x$ in the context $\Gamma$.  When an expression
is a variable \code{(VarMa\ \_\ $\Gamma$\ i)}, the denoted value is the
corresponding variable in the context, i.e., \code{(index\ i\ $\Gamma$)}.  The
expression \code{(ConstMa\ \_\ $\Gamma$\ k)} denotes the constant $k$ in any context $\Gamma$. Finally, if $e_x$ is an expression encoding the concrete value $x$, and
$e_y$ is an expression encoding the concrete value $y$, then the expression
\code{PlusMa\ $e_x$\ $e_y$} denotes the concrete value $(x + y)$. Because the
reflected terms embed their corresponding inputs, they are guaranteed to be
sound representations.

@@

We will explain how to implement the tactic for \code{Group} specifically,
and the other structures are implemented similarly.
The equivalence we are trying to prove is $x \simeq y$, where $x$ and $y$ are
elements of the type $c$, which implements a group. The reflected term for $x$
is denoted $e_x$, and has type $\code{ExprG}\ p\
\Gamma\ x$, which means that $e_x$ is guaranteed to be the encoding of $x$.  
Similarly,
$y$, is encoded by $e_y$, and its type is indexed by the
value $y$.  Evaluating \code{norm} on $e_x$ produces the
normal form $e_{x'}$ of type $\code{ExprG}\ p\ \Gamma\ x'$ and a proof $p_x$ of
$x \simeq x'$.
Similarly, for $e_y$, it produces
the normal form $e_{y'}$ of type $\code{ExprG}\ p\ \Gamma\ y'$ and a proof
$p_y$ of $y \simeq y'$.  It now suffices to compare the normal forms $e_{x'}$ and $e_{y'}$ using a syntactic equivalence test \code{ExprG\_eq}, because once everything is in normal form, being equivalent is just a matter of being syntactically equivalent.

\begin{lstlisting}
exprG_eq : {c:Type} $\rightarrow$ {n:Nat} $\rightarrow$ (p:Group c) $\rightarrow$ ($\Gamma$:Vect n c) 
           $\rightarrow$ {x' : c} $\rightarrow$ {y' : c} 
           $\rightarrow$ (ex' : ExprG p $\Gamma$ x') $\rightarrow$ (ey' : ExprG p $\Gamma$ y') 
           $\rightarrow$ Maybe(x' $\simeq$ y')
\end{lstlisting}
This syntactical equivalence test checks if the two input terms $ex'$ and $ey'$ are \emph{syntactically} the same, and if they do, it directly builds a proof of equivalence between their indices $x' \simeq y'$, which is what we need, because we can use it 
with the two equivalences $x \simeq x'$ and $y\simeq y'$ that we already have, in order to get the desired proof of $x \simeq y$ with \code{eq\_preserves\_eq}. We put all of this together in a function \code{buildProofGroup}:

@@

Each algebraic structure is equipped with a function for reducing reflected terms to their normal form.
The algebraic theories which concern us admit a canonical representative\footnote{It only holds for ``pure'' algebraic structures, ie, in the absence of additional axioms}  for
any element, a property which we use to decide equalities. Without this
property, it would be more complicated to decide equivalence without
brute-forcing a series of rewritings, that would have no termination guarantee.
